Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-02-04 analysis/paper/paper.Rmd
editing the sample section

Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP., what their proportions are in local culinary practices during the early time are unclear. For example, plant remains from Guijiabao show that millet is the major crop, which is at odds with the predominance of rice based on phytolith analysis. For Gaoshan and neighboring sites in Sichuan and Yunnan, such as Baodun, Henglanshan, and Baiyangcun, it appears that rice accounts for the vast majority of plaint remain [@Hao2022; @Guedes2013]. The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season. Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP.) and the historical period (1000 BP.). The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022]. Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them has incised cord-marked decoration. Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a].

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Evershed1994].

We cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS. Internal standards were added before and after the total lipid extraction.

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of samples, suggesting a good preservation of lipids. Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millets), or new species (eg. pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020]. It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Due to our small samples, it is hard to determine the role of rice/millet during the Neolithic period. Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

>>>>

Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP., what their proportions are in local culinary practices during the early time are unclear. For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in neighboring sites, such as Gaoshan and Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013]. The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks. This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.

@@

## Samples
We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX). Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season. Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP.) and the historical period (1000 BP.). The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022]. Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them have incised cord-marked decoration. Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a]. This research also attempts to confirm that those ceramics were cooking pots used in the daily life of the local residents.

## Methods
For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For the sample processing, we cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS. Internal standards were added before and after the total lipid extraction.

@@

One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of ceramic or a different use. Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.

@@

<!--discussion about possible plant origins-->
One aim of our study is to explore whether new crops, such as C3 plants (eg. rice), C4 plants (eg. millets), or new species (eg. pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition. We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020]. It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016]. However, the seeds of wild weeds only account for a small part of plant remains. Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022]. Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
